A Neural Network Alternative to Convolutive Audio Models for Source Separation

### 2017-06-13 paper/paper.tex
I have made changes in paper.tex

Non-negative matrix factorization~(NMF) of magnitude-spectrograms has been a very popular method for modeling sources for supervised source separation applications~\cite{smaragdis2014static, virtanen2015compositional}. NMF factorizes a matrix of non-negative elements~$\mathbf{X}\in \mathbb{R}_{M\times N}^{\geq0}$ as a product of the basis matrix~$\mathbf{W}$ and the activation matrix~$\mathbf{H}$. The notation $\mathbb{R}_{M \times N}^{\geq0}$ represents the set of matrices of non-negative elements of size $M \times N$. In this factorization, the basis matrix $\mathbf{W} \in \mathbb{R}_{M \times r}^{\geq{0}}$, the activation matrix $\mathbf{H} \in \mathbb{R}_{r \times N}^{\geq{0}}$ and $r$ represents the rank of the decomposition. In the case of audio signals, we apply NMF on audio spectrograms. In the audio setting, the columns of $\mathbf{W}$ begin to act as representative basis vectors for the source. The rows of $\mathbf{H}$ indicate the activity of these basis vectors in time. \textbf{C: I am not sure how smooth the transition to neural nets here.} As shown in~~\cite{smaragdis2017aneural}, the notion of non-negative audio modeling can be easily generalized by interpreting NMF as a neural network. We can interpret NMF as a non-negative auto-encoder in the following manner,
\begin{align}
    \text{$1^{\text{st}}$ layer:~}\mathbf{H} &= g(\mathbf{W^{\ddagger}} \cdot \mathbf{X}) \nonumber \\
    \text{$2^{\text{nd}}$ layer:~}\mathbf{X} &= g(\mathbf{W} \cdot \mathbf{H})
    \label{eq:nmfae}
\end{align}

Here, $\mathbf{X}$ represents the input spectrogram, $\mathbf{W^{\ddagger}}$ represents a form of pseudo-inverse of $\mathbf{W}$ and $g(.):\mathbf{R}\rightarrow \mathbf{R}^{\geq0}$ is an element-wise function that maps a real number to the space of positive real numbers. As before, the columns of $\mathbf{W}$ act as representative basis vectors and the corresponding rows of $\mathbf{H}$ indicate their respective activations. Although non-negativity of the network-parameters (models) is not explicitly guaranteed in this formulation, applying a suitable sparsity constraint allows the network to learn suitable non-negative models. Additionally, this interpretation enables a pathway to propose multi-layer extensions by exploiting the wealth of available neural net architectures that could potentially lead to superior separation performance. \\

Spectrograms of speech and audio signals incorporate temporal dependencies that span multiple time frames. However, NMF and its neural network equivalent are unable to explicitly utilize these cross-frame patterns available in a spectrogram. To alleviate this drawback, Smaragdis~\cite{smaragdis2007convolutive} proposed a convolutive version to NMF~(conv-NMF) that allows spectro-temporal patterns as representative basis elements. In this paper, we develop a neural network alternative to such convolutive audio models for supervised source separation. In doing so, we solve two fundamental problems associated with this task. The first step is to develop a suitable neural network architecture to learn convolutive audio models in an adaptive manner. Utilizing the models to separate a source from a given mixture forms the second step. \\

Developing a neural network alternative to conv-NMF also results in a significant advantage in terms of model re-usability. Several neural network architectures have been recently proposed for supervised source separation~\cite{grais2017single, venkataramani2017end}. These networks are trained to learn discriminative audio models by training them on mixture spectrograms. Thus, the networks learn a decomposition that works only for a specific source-interference pair. Developing an alternative to conv-NMF allows us to learn convolutive audio-models by training neural networks on clean source spectrograms. These models can be easily re-used irrespective of the interfering component. At the same time, it is also easy to propose extensions to the models by using different and exotic neural network layers. \\

@@

\textbf{C: Impulse response concept is for linear systems. I am not too sure about using it in this context.} The decoder CNN can be interpreted as a finite impulse response~(FIR) filter. The encoder attempts to approximate the inverse of the decoder FIR filter as an FIR filter. From our knowledge of signal processing, it can be shown that the inverse filter of an FIR filter has an infinite impulse response~(IIR)~\cite{smith2011spectral}\footnote{\text{http://ccrma.stanford.edu/\~{}jos/fp/Inverse\_Filters.html}}. Thus, approximating the encoder as an FIR filter or a CNN will hinder the decoder from learning patterns that are invertible only by an IIR filter. Therefore, we explore using a Recurrent Neural Network~(RNN) to get an IIR encoder. \\

\section{Using a recurrent filter in the encoder}
\label{sec:rnncnn}
In this section, the goal is to construct a recurrent encoder analogous to the convolutive encoder we discussed in the previous section. We will refer to this auto-encoder as a Recurrent-Convolutive Auto-encoder~(RCAE). The potential gain of using a recurrent encoder over a finite length convolutive encoder is due to the fact that a recurrent filter in theory can capture arbitrarily long temporal dependencies. From a signal processing point of view, the motivation for going with a recurrent filter is that, the inverse of an finite length filter (the convolutive basis in the encoder) is given by a recurrent filter. We explore the idea, and obtain source separation results on the same dataset. \\

>>>>

Non-negative matrix factorization~(NMF) of magnitude-spectrograms has been a very popular method for modeling sources for supervised source separation applications~\cite{smaragdis2014static, virtanen2015compositional}. NMF factorizes a matrix of non-negative elements~$\mathbf{X}\in \mathbb{R}_{M\times N}^{\geq0}$ as a product of the basis matrix~$\mathbf{W}\in \mathbb{R}_{M \times r}^{\geq{0}}$ and the activation matrix~$\mathbf{H}\in \mathbb{R}_{r \times N}^{\geq{0}}$. The notation $\mathbb{R}_{M \times N}^{\geq0}$ represents the set of matrices of non-negative elements of size $M \times N$, and $r$ represents the rank of the decomposition. In the case of audio signals, NMF is applied on audio spectrograms, where the columns of $\mathbf{W}$ act as representative basis vectors for the source. The rows of $\mathbf{H}$ indicate the activity of these basis vectors in time.  \\

As shown in~~\cite{smaragdis2017aneural}, the notion of non-negative audio modeling can be easily generalized by interpreting NMF as a neural network. We can interpret NMF as a non-negative auto-encoder in the following manner,
\begin{align}
    \text{$1^{\text{st}}$ layer:~ (Encoder)~}\mathbf{H} &= g(\mathbf{W^{\ddagger}} \cdot \mathbf{X}) \nonumber \\
    \text{$2^{\text{nd}}$ layer:~ (Decoder)~}\mathbf{X} &= g(\mathbf{W} \cdot \mathbf{H})
    \label{eq:nmfae}
\end{align}

Here, $\mathbf{X}$ represents the input spectrogram, $\mathbf{W^{\ddagger}}$ represents a form of pseudo-inverse of $\mathbf{W}$ and $g(.):\mathbf{R}\rightarrow \mathbf{R}^{\geq0}$ is an element-wise function that maps a real number to the space of positive real numbers. As before, the columns of $\mathbf{W}$ act as representative basis vectors and the corresponding rows of $\mathbf{H}$ indicate their respective activations. Although non-negativity of the network-parameters (models) is not explicitly guaranteed in this formulation, applying a suitable sparsity constraint allows the network to learn suitable non-negative models. Additionally, this interpretation enables a pathway to propose variants to this basic autoencoder stucture by exploiting the wealth of available neural net architectures that potentially lead to superior separation performance. The main alternative we explore in this paper is the convolutive architecture, which is based on convolutive NMF \cite{smaragdis2007convolutive}.  \\

Spectrograms of speech and audio signals incorporate temporal dependencies that span multiple time frames. However, NMF and its neural network equivalent are unable to explicitly utilize these cross-frame patterns available in a spectrogram. To alleviate this drawback, Smaragdis~\cite{smaragdis2007convolutive} proposed a convolutive version to NMF~(conv-NMF) that allows spectro-temporal patterns as representative basis elements. In this paper, we develop a neural network alternative to such convolutive audio models for supervised source separation. \textbf{C:What does this following three sentences mean?}In doing so, we solve two fundamental problems associated with this task. The first step is to develop a suitable neural network architecture to learn convolutive audio models in an adaptive manner. Utilizing the models to separate a source from a given mixture forms the second step. \\

Several neural network architectures have been recently proposed for supervised source separation~\cite{grais2017single, venkataramani2017end}, which are discriminatively trained on audio mixtures. Although discriminative training of a source separation network results in good seperation performance, the network is restricted to work on a particular type of mixture. The convolutive architecture that we mainly discuss in this paper is generatively trained, and therefore not restricted to the types of mixtures that is used in training. By the virtue of flexibility of neural networks, we also propose a variant where recurrent neural network (RNN) is used. \\

@@

Note that the encoder attempts to approximate the inverse of the decoder. From our knowledge of linear filtering in signal processing, we know that the inverse of a finite length filter is given by a recursive filter. ~\cite{smith2011spectral}\footnote{\text{http://ccrma.stanford.edu/\~{}jos/fp/Inverse\_Filters.html}}. Following this analogy, in the next section we explore using a recurrent neural network in the encoder.

\section{Using a recurrent filter in the encoder}
\label{sec:rnncnn}
In this section, the goal is to construct a recurrent encoder analogous to the convolutive encoder we discussed in the previous section. We will refer to this auto-encoder as a Recurrent-Convolutive Auto-encoder~(RCAE). The potential gain of using a recurrent encoder over a finite length convolutive encoder is due to the fact that a recurrent filter in theory can capture arbitrarily long temporal dependencies. From a signal processing point of view the motivation for going with a recurrent filter is that, the inverse of an finite length filter (the convolutive basis in the encoder) is given by a recurrent filter. \\
